GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-03-11 manuscript.tex
Updates in GSP 1.1

\section{Introduction}\label{sec:introduction}
The GeoSPARQL standard, first issues in 2021 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
is one of, if not the most\footnote{It is hard to calculate use but references to GeoSPARQL in papers and other 
well-known standards, such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used 
\textit{Sematic Web} ontologies for representing spatial data.

@@

Specifically ruled out of scope was:

\begin{itemize}
    \item[$\ast$] investigation of property graphs
\end{itemize}

This exclusion was due to the recent (last several years) discussion in the OGC and elsewere about property graphs. The SWG 
proponents felt that while property graphs might be important for future \textit{Semantic Web} spatial data systems, there was
more than enough work scoped for initial SWG investigations already so as to exclude this are of investigationfor initial 
GeoSPARQL revisions.

@@

\begin{itemize}
    \item[$\ast$] GeoSPARQL 1.1: extensions that are fully compatable with GeoSPARQL 1.0
    \item[$\ast$] GeoSPARQL 1.2: fully or mostly compatable extensions but which are larger additions to the standard's conceptual coverage
    \item[$\ast$] GeoSPARQL 2.0: a future GeoSPARQL that might be quite different and partly incompatable with GeoSPARQL 1.0
\end{itemize}

The reason for expecting a future, incompatable, GeoSPARQL 2.0 is that early SWG attendees thought spatio-temporal relations
and fundamental ontology elements in GeoSPARQL either could or should be remodelled which might break the current, familiar, 
\texttt{Feature}/\texttt{Geometry} class relations. Details of these potential changes haven't been fully expounded, at the 
time of this paper, however initial SWG attendees' intuition is that a future GeoSPARQL 2.0 might generalise spatial concepts and
move away from only, or primarily, \textit{geo}spatial, or perhaps focus not just on \texttt{Feature}/\texttt{Geometry} relations
but look to generalised mecahnisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
\textit{temporality} too).

@@

\section{New features in GeoSPARQL 1.1}\label{sec:newfeatures}
One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}\cite{atkinson_profiles_2020}. 
The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
of which were easy to discover and, as a result, some GeoSPARQL users were unaware of some of the resources and some
resources were accidentally duplicated or partly re-implemented.

The profile declaration will be published by the OGC as a stand-alone resource sometime in mid-2021 along with some 
updated GeoSPARQL 1.0 resources. Currently, all of the elements of GeoSPARQL 1.0, including the profile declaration, 
can be found within the SWG's working online repository \footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.

% GeoSPARQL 1.1 includes non-breaking changes to the GeoSPARQL 1.0 standard.

\subsection{New geometry literals}\label{sec:literals}
GeoSPARQL 1.1 introduces three new literal types: A GeoJSON\cite{butler2016geojson} literal, a KML\cite{nolan2014keyhole} 
literal, and a DGGS\cite{sahr1998discrete} WKT Literal.

\small
\begin{lstlisting}[caption=GeoJSON literal example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily]
"{\"type\":\"Point\", \"coordinates\":[-83.38,33.95]}"
^^<http://www.opengis.net/ont/geosparql#geoJSONLiteral>
\end{lstlisting}
\normalsize
Since the GeoJSON and KML formats are restricted to be represented in the WGS84 SRS only, these restrictions 
also apply to GeoJSON and KML literals. The DGGS WKT literal  
\subsection{Spatial aggregate functions}\label{sec:spatialaggregate}
GeoSPARQL 1.1 includes new spatial aggregate functions, which allow for the quick aggregation of different geometry types. 
While spatial aggregate functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
at the time of defining the GeoSPARQL 1.0 standard, aggregate functions had not yet been introduced into the SPARQL standard 
since SPARQL 1.1 \cite{w3c_sparql_working_group_sparql_2013} was released about one year later. Spatial aggregate functions 
similar to traditional aggregate functions such as AVG,MAX, or MIN allow to aggregate results of geometry queries, e.g., to 
create the union out of a set of given geometry literal results. While calculating these aggregates may also be possible 
outside of the semantic database, the inclusion of the functions provides distinct advantages:
\begin{enumerate}
    \item No client-side library is needed to create an aggregated geometry result
    \item Fewer and more appropriate results can be returned using a spatial aggregate function (e.g., a single union geometry vs. a set of geometries for which to create union externally)
    \item Spatial aggregates from different SPARQL endpoints can easily be calculated using federated queries 
\end{enumerate}
GeoSPARQL 1.1 defines the aggregate functions \emph{geosf:boundingCircle}, calculating a bounding circle around a set of 
geometry, \emph{geosf:centroid} calculating the centroid of the set of geometries, \emph{geosf:concatLines} concatenating 
a set of linestrings that overlap, \emph{geosf:concaveHull} calculating the concave hull of a set of geometries and\emph{geosf:union}, 
calculating the union of a set of geometreis, in addition to the \emph{geosf:envelope} and \emph{geosf:convexHull} functions 
already defined in the GeoSPARQL 1.0 standard. Since SPARQL 1.1 these fucntions were also usable as aggregate functions.

@@

\subsection{Extensions of the GeoSPARQL ontology}\label{sec:ontologyextensions}
GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{geo:SpatialMeasure}. 
This class represents a measurement such as a volume, length, or area associated with a measurement amount and a measurement 
unit. It acts as a range of three newly defined properties \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasArea}, 
\href{http://www.opengis.net/ont/geosparql#hasLength}{geo:hasLength}, \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasVolume} 
which make these attributes of a geometry better accessible using SPARQL. Finally, GeoSPARQL 1.1 adds the 
\href{http://www.opengis.net/ont/geosparql#inSRS}{geo:inSRS} property, which allows querying the SRS of a given geometry 
without the need to analyze geometry literal serializations. The property allows for the definition of a SRS as a URI. It 
paves the way for a future definition of SRS fully in RDF as anticipated for GeoSPARQL 2.0.

>>>>

\section{Introduction}\label{sec:introduction}
The GeoSPARQL standard, first issues in 2021 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
is one of, if not the most\footnote{It's hard to calculate use but references to GeoSPARQL other well-known standards, 
such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used \textit{Sematic Web} ontologies for 
representing spatial data.

@@

Specifically ruled out of scope was any investigation of property graphs. Recent (last several years) discussion in the OGC and 
elsewere about property graphs moticated a consideration of them, hoever the SWG proponents felt that while property graphs might 
be important for future \textit{Semantic Web} spatial data systems, there was more than enough work scoped for initial SWG work
(several revisions of the standard) to initially exclude this are of investigation.

@@

\begin{itemize}
    \item[$\ast$] \textbf{1.1}: extensions that are fully compatable with GeoSPARQL 1.0
    \item[$\ast$] \textbf{1.2}: fully or mostly compatable extensions but which are larger additions to the standard's conceptual coverage
    \item[$\ast$] \textbf{2.0}: a future GeoSPARQL that might be quite different and partly incompatable with GeoSPARQL 1.0
\end{itemize}

The reason for expecting a future, incompatable, GeoSPARQL 2.0 is that early SWG attendees thought spatio-temporal relations
and fundamental ontology elements in GeoSPARQL either could or should be remodelled which might break the current, familiar, 
\texttt{Feature}/\texttt{Geometry} class relations. Details of these potential changes haven't been fully expounded, at the 
time of this paper, however initial SWG attendees' intuition is that a future GeoSPARQL 2.0 might generalise spatial concepts and
move away from only, or primarily, \textit{geo}spatial, or perhaps focus not just on \texttt{Feature}/\texttt{Geometry} relations
but look to generalised mechanisms for describing dimensions of features of which \textit{geometry} is one of many (and perhaps)
\textit{temporality} too).

@@

\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
So far (as of June, 2020) the GeoSPARQL SWG has triarged changed requests into GeoSPARQL 1.1, 1.2 \$ 2.0 releases and has addressed
many 1.1 requests. Here we report only on those 1.1 requests. This section describes some of the areas updated in GeoSAPRQL 1.1
and the following section foreshadows likely 1.2 and 2.0 updates.

\subsection{Profile Declaration}\label{sec:profiledec}
One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}\cite{atkinson_profiles_2020}. 
The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
of which were easy to discover and, as a result, some GeoSPARQL users were unaware of some of the resources and some
resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
as being the \textit{best practice} way for it to deliver multi-part standards.

The profile declaration for GeosPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
updated GeoSPARQL 1.0 resources. Currently, all of the elements of GeoSPARQL 1.0, including the profile declaration, 
can be found within the SWG's working online repository \footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.
The profile declaration for GeosPARQL 1.1 will be published at the same time as all, or most, of the 1.1 releases' updated
resources, currently expected in mid-2021.

\subsection{New geometry literals}\label{sec:newliterals}
GeoSPARQL 1.1 introduces three new litergeometry serializations:

\begin{enumerate}
    \item GeoJSON (Geo- JavaScript Object Notation)\cite{butler2016geojson}
    \item KML (Keyhole Markup Language)\cite{nolan2014keyhole} 
    \item DGGS (Discrete Global Grid System)\cite{sahr1998discrete}
\end{enumerate}

The first two of these are ``obvious'' in the sence that that they have been expected/requested by the SDWWG and many users
of GeoSPARQL for a long time, due to those format's popularity. an example of a GeoJSON geometry serialization is given below.

\small
\begin{lstlisting}[caption=GeoJSON geometry serialization example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily]
"""{"type":"Point", "coordinates":[-83.38,33.95]}"""
^^<http://www.opengis.net/ont/geosparql#geoJSONLiteral>
\end{lstlisting}
\normalsize
Since the GeoJSON and KML formats are restricted to be represented in the WGS84 SRS only, these restrictions 
also apply to GeoJSON and KML literals. The DGGS WKT literal

The third new serialization format is more forward-looking in that it is not dirven by user demand but by predected demand.
DGGS does not have a single, concrete format standard as the others do, not is it ever likely to - different DGGSes will 
likely implement very different data formats - so GeoSAPRQL 1.1 makes generalized provisions for DGGS serializations but 
presents no detailed requirements for them, only stating that the specific DGGS must be identified.

\subsection{New spatial functions}\label{sec:newfunctions}
GeoSPARQL 1.1 includes new spatial aggregate functions, which allow for the quick aggregation of different geometry types. 
While spatial aggregate functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
at the time of defining the GeoSPARQL 1.0 standard, aggregate functions had not yet been introduced into the SPARQL standard 
since SPARQL 1.1 \cite{w3c_sparql_working_group_sparql_2013} was released about one year later. Spatial aggregate functions 
similar to traditional aggregate functions such as AVG,MAX, or MIN allow to aggregate results of geometry queries, e.g., to 
create the union out of a set of given geometry literal results. While calculating these aggregates may also be possible 
outside of the semantic database, the inclusion of the functions provides distinct advantages:

\begin{enumerate}
    \item No client-side library is needed to create an aggregated geometry result
    \item Fewer and more appropriate results can be returned using a spatial aggregate function (e.g., a single union geometry vs. a set of geometries for which to create union externally)
    \item Spatial aggregates from different SPARQL endpoints can easily be calculated using federated queries 
\end{enumerate}

GeoSPARQL 1.1 defines the aggregate functions \emph{geosf:boundingCircle}, calculating a bounding circle around a set of 
geometry, \emph{geosf:centroid} calculating the centroid of the set of geometries, \emph{geosf:concatLines} concatenating 
a set of linestrings that overlap, \emph{geosf:concaveHull} calculating the concave hull of a set of geometries and\emph{geosf:union}, 
calculating the union of a set of geometreis, in addition to the \emph{geosf:envelope} and \emph{geosf:convexHull} functions 
already defined in the GeoSPARQL 1.0 standard. Since SPARQL 1.1 these fucntions were also usable as aggregate functions.

@@

\subsection{Ontology extensions}\label{sec:ontexts}
GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{geo:SpatialMeasure}. 
This class represents a measurement such as a volume, length, or area associated with a measurement amount and a measurement 
unit. It acts as a range of three newly defined properties \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasArea}, 
\href{http://www.opengis.net/ont/geosparql#hasLength}{geo:hasLength}, \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasVolume} 
which make these attributes of a geometry better accessible using SPARQL. Finally, GeoSPARQL 1.1 adds the 
\href{http://www.opengis.net/ont/geosparql#inSRS}{geo:inSRS} property, which allows querying the SRS of a given geometry 
without the need to analyze geometry literal serializations. The property allows for the definition of a SRS as a URI. It 
paves the way for a future definition of SRS fully in RDF as anticipated for GeoSPARQL 2.0.
